---
title: "F-tests"
author: "George Melrose"
date: "12/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### F-tests

An F-test tests whether two population variances are equal.

H0 - Population variances are equal

H1 - Population variances are not equal

To do an F-test, one can use the *var.test()* function in one of two ways:

* var.test(x,y, alternative = "two.sided")

* var.test(values ~ groups, data, alternative = "two.sided")

*Alternative* indicates the alternative hypothesis.

```{r}
#define groups#
x <- c(76,65,78,69,78,69)
y <- c(75,87,59,60,63,65)

#perform F-test to determine if variances are equal#
var.test(x,y)
```
The p-value is 0.260685 - less than 0.05 so the H0 is rejected. There's enough evidence to suggest the population variances aren't equal.

```{r}
#defining groups#
data <- data.frame(values=c(20,20,27,31,25,19,23,26,35,31,30,28,19,20),group=rep(c('A','B'), each=7))

#performing an F-test#
var.test(values~group, data = data)
```
The p-values is way above .05 and so the H0 is accepted - population variances are equal.

Two questions usually are answered by the F-test -

* Do two samples come from populations with equal variances?

* Does a new treatment or process reduce the variability of some current treatment or process?
