---
title: "Fisher's_Exact_Test"
author: "George Melrose"
date: "31/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fisher's Exact Test 

Fisher's exact test is used instead of a chi-squared test, with 2x2 tables, when samples sizes are too small. Firstly, a 2x2 matrix must be generated. 
```{r}
data = matrix(c(3,6,7,8), nrow=2)
print(data)
```
To conduct Fisher's Exact test, one simply uses the *fisher.test()* function.

```{r}
fisher.test(data)
```
In Fisher's exact test, the H0 is that the 2 columns are independent/ the odds ratio (OR) =1. To determine this, the p-value is useful.As the p-value is way above the normal 0.05 threshold,at 0.6785, there isn't enough evidence to reject the null hypothesis. There isn't any statistically significant difference between the two columns. 

Additionally, the OR of 0.585 indicates that there's a low odds of association between the two columns. The 95% confidence interval (0.06785812, 4.14635455) contains 1 within the ratio, confirming that the OR isn't significantly different to 1. 


