---
title: "Pooled SD"
author: "George Melrose"
date: "12/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pooled Standard Deviation 

Pooled SD is the weighted average of SDs from two or more independent groups. It most often crops up in the two-samples t-test, when ones determines whether or not the means of two populations are equal.

The formula for pooled SD is - sqrt((n1-1)sqr(s1) + (n2-1)sqr(s2)/ (n1+n2-2))

where:

* n1,n2: Sample sizes for group 1 and group 2 respectively.

* s1,s2: SD for group 1 and group 2 respectively.

### Manual calculation
```{r}
#define samples#
data1 <- c(6,7,5,1,9,98,1,14,106,2)
data2 <- c(5,6,1,7,15,145,158,78,96,46,32,4)

#find SD of each sample#
s1 <- sd(data1)
s2 <- sd(data2)

#find size of each samples#
n1 <- length(data1)
n2 <- length(data2)

#pooled SD#
pooled <- sqrt(((n1-1)*s1^2 + (n2-1)*s2^2) / (n1+n2-2))
print(pooled)
```

### Using a function

```{r}
library(effectsize)

sd_pooled(data1, data2)
```

