---
title: "Weighted Standard Deviation"
author: "George Melrose"
date: "12/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Weighted Standard Deviation

Weighted standard deviation is used to measure the dispersion of dataset values, when some values in a dataset have much higher weights than others.The most straightforward way to do this is to use the *wt.var()* function from the *Hmisc* package.

When talking about a weight variable, it means just that - observations having more weight have more influence in analysis than oversvations with smaller weights. An unweighted analysis means a weighted analysis where all weights are 1. A great explanation of statistics 'weights' is provided here - https://blogs.sas.com/content/iml/2017/10/02/weight-variables-in-statistics-sas.html

```{r}
#Data values#
x <- c(5, 56, 7, 8 ,19, 107, 68)

#Weights# 
wt <- c(.5,1,2,3,2,1,.4)

library(Hmisc)
#Weighted variance#
weighted_var <- wtd.var(x, wt)

#weighted variance SD#
sqrt(weighted_var)
```

### Weighted SD for one column of a DF

```{r}
library(Hmisc)

#dataframe#
df <- data.frame(team=c('Man U','Man City','Liverpool','Newcastle','Boro','Chelsea','Leeds'),wins = c(2,19,3,5,3,11,10), points=c(7,45,10,12,9,28,20))

#weights#
wt <- c(1,1,1.5,0.5,2,1,0.8)

#weighted SD of points#
sqrt(wtd.var(df$points, wt))
```
### Weighted SD of multiple columns of a DF

One can use the *sapply()* function to calculate weighted SD for multiple columns in a DF. Sapply() is a built-in function that applies a function to all input elements be they lists, vectors, or DFs. Sapply() returns a vector or matrix.

```{r}
sapply(df[c('wins', 'points')], function(x) sqrt(wtd.var(x,wt)))
```



